Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-03 appendix.tex
be more explicit

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
%EM We don't think they occur.
Discontinuities occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we are at a point of discontinuity we maximize over ancestral state splits, ensuring the upper-semicontinuous property of this mapping.
This shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Perhaps we can say something about how this is not a strong assumption.
%EM see below "By the assumption of"
Assume $x_1,y_1,x_2 < 1$.
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
Assuming $x_1,y_1,x_2 < 1$ precludes all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
Moreover, we have
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0. Maybe I'm being too careful? I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
The remaining variables follow similar arguments.

@@

%EM Also here.
Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
We now prove the main result.

@@

Thus, if $\ell'_{L}(1) > 0$ then $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.

%EM Topic sentence here please.
The conditions on $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$ are straightforward, the one on the derivative being
\[
\ell'_{L}(1) =  -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
\]
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
\end{proof}

>>>>

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
%EM Do discontinuities occur? In my mind we are taking the maximum of continuous functions, which is continuous.
%EM We don't think they occur.
%das: done
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Perhaps we can say something about how this is not a strong assumption.
%EM see below "By the assumption of"
%das: moved below
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
In other words, we make the assumption that $x_1,y_1,x_2 < 1$, precluding all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
This assumption is reasonable as the likelihood is never maximized at these values where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$.
Moreover, we have
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0. Maybe I'm being too careful? I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
The remaining variables follow similar arguments.

@@

%EM Also here.
%das: reworded
Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ does not contradict with the statement of Lemma~\ref{lemma:bounds}.
We now prove the main result.

@@

Thus, if $\ell'_{L}(1) > 0$, since $\ell'_{L}(w)$ is decreasing on $(0,1]$ then $\ell'_{L}(w) > 0$ for all $w\in (0,1]$.
Since $\ell'(w) > \ell'_{L}(w)$ for $w\in(0,1]$, this implies $\ell'(w) > 0$, which further implies $\ell(w; \boldsymbol\xi_0)$ is increasing on the entire interval $(0,1]$, including at the boundary $w=1$, and $\hat{w} \equiv 1$.

%EM Topic sentence here please.
%das: made more explicit and introduced more gently
We now construct the conditions on $x^*$ and $y^*$ that are required for $\hat{w}\equiv 1$:
\[
C1-C3: b_{2}^U, b_{1}^U, b_{12}^U < 1,
\]
\[
C4: -\frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} + \frac{1}{2} > 0.
\]
Conditions $C1-C3$ are needed so that $\ell'_{L}(w)$ is a decreasing function on $(0,1]$, while condition $C4$ is the one on $\ell'_{L}(1) > 0$, i.e., so that $\ell'(w)$ is positive on $(0,1]$.
The set of $x^*,y^*$ satisfying conditions $C1-C4$ as well as those for $\emptyset$ being the most likely ancestral state split is given in Fig.~\ref{fig:incons-analytic}, giving the region of generating parameters where $\hat{w}\equiv 1$.
\end{proof}
